---
title: "Activity Recognition: Prediction Assignment Writeup"
author: Pratheep Pakkianathan
output: html_document
---

###Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Description of the experiment and data

From: http://groupware.les.inf.puc-rio.br/har

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg).

###Overview of analysis
Since the testing sample does not have the dependant variable, the training sample itself will be partitioned into training and validation sub samples.  The algorithm will learn on the training subsample and then tested against the validation subsample, before being applied on the testing sample, after assessment of the out of sample error.

To validate and cross check the prediction two different predictions algorithms are used.  C.50 will be applied first, followed by randomForrest for comparison of results.

###Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 


###Data procssing.

```{r, message=F, warning=F}
#Load required libraries
library(caret)
library(kernlab)
library(C50)
library(randomForest)
```


```{r}
set.seed(12345)
```

####Data loading and cleaning


```{r}
if (! file.exists('./pml-training.csv')) {
    download.file('http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv', destfile = './pml-training.csv')
}
if (! file.exists('./pml-testing.csv')) {
    download.file('http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv', destfile = './pml-testing.csv')
}
```


Both test and training datasets were  were downoaded to working directory and scanned for NA values. The following were identified as na.strings: "NA","#DIV/0!",""

The first seven columns are not needed for prediction. They are removed.
Any columns with near zero variance are also removed.


```{r}
mltraining <- read.csv("./pml-training.csv", na.strings=c("NA","#DIV/0!",""))
mltesting <- read.csv("./pml-testing.csv", na.strings=c("NA","#DIV/0!",""))


mltraining <- mltraining[, -c(1:7)]
mltesting <- mltesting[, -c(1:7)]

# Remove rezo variance columns
nearzerocols <- nearZeroVar(mltraining)
mltraining <- mltraining[, -nearzerocols]
mltesting <- mltesting[, -nearzerocols]

```

Keep variables with less than 40% of missing data. 

```{r}
good_data_columns <- apply(mltraining,2, function(x) {  sum(is.na(x)) / length(x) < 0.4 }  )
mltraining <- mltraining[, good_data_columns]
mltesting <- mltesting[, good_data_columns]

#Data after cleaned up
dim(mltraining)
dim(mltesting)
table(mltraining$classe)
```


####Data partitioning

The training data is partitioned into two subsamples: One  is used for training (60%) and the other used for validation and fine tuning of the modal


```{r}
sub_training <- createDataPartition( mltraining$classe, p =0.6, list = F )
sub_training_sample <- mltraining[sub_training, ]
sub_validation_sample <- mltraining[-sub_training, ]
dim(sub_training_sample)
dim(sub_validation_sample)
```

####Prediction

C50 decision tree algorithm will be used to train, with 10 trials. The main advantage is interpretability and the algorithm is fairly efficient. Later, the predicted results will be cross checked against random Forest.

summary(c50fit) will print the entire C50 tree with additional information. Since the output is too long for this report, it's not shown here. 


```{r}
sub_training_sample_c50 <- sub_training_sample[, -53]
c50fit <- C5.0(sub_training_sample_c50, sub_training_sample$classe, trials = 10)
#Summary output will be too long
c50fit
```

Make a prediction on the training data, followed by the validation sample. Comparing the two confusion matrices gives an indication of  model performance 

```{r}
c50trainPred <- predict(c50fit, sub_training_sample)
confusionMatrix(c50trainPred, sub_training_sample$classe)
```

Insample accurcy is 1. Sample error is 0

```{r}
c50ValidationPred <- predict(c50fit, sub_validation_sample)
confusionMatrix(c50ValidationPred, sub_validation_sample$classe)
```

####Sample error
Out of sample error is very small, 0.01. ( 1 - Accuracy,0.9892)  


The model will now be used on the testing dataset to predict the outcome.

```{r}
c50TestPred <- predict(c50fit, mltesting)
c50TestPred 
```

To double check the results, a random forest will be built against the training subset and applied on the test sample. 

```{r}
rffit <- randomForest(classe ~ ., data = sub_training_sample,  ntrees = 10)
#rfValidationPred <- predict(rffit, sub_validation_sample)
#confusionMatrix(rfValidationPred, sub_validation_sample$classe)
rffit.pred <- predict( rffit ,  mltesting)
rffit.pred
```

###Conclusion

Final predcited result is
```{r}
c50TestPred
```

Both, random Forest and C50 predicted the same result, an indication the prediction was highly accurate.  
